Efficient ancestry and mutation simulation with msprime 1.0

### 2021-11-14 response-to-reviewers.tex

\begin{document}

\section*{Response to the reviewers}
% General intro text goes here
We thank the reviewers for their close reading of our manuscript and
insightful comments. In the following we address the points raised
in turn.

% Let's start point-by-point with Reviewer 1
\reviewersection

% Point one description
\begin{point}
Simulation is a vital part of population genetics, and msprime has quickly
become a crucial tool for many researchers. msprime was first presented in a
publication in 2016, where the main focus was the use of "sparse trees" and
"coalescence records" to greatly increase the efficiency and practicality of
large coalescent simulations. In this new manuscript, the authors introduce a
new version of the software, with this manuscript ""focusing on the aspects
that are either new for this version, or in which our approach differs
significantly from classical methods."

Since 2016, msprime has gained popularity not only due to its computational
efficiency, but also for its software development style with well-documented
features, development guidelines, and ever-growing capabilities. The extensive
author list shows how the community around msprime and associated programs has
grown over the last five years. The current manuscript has a huge number of
citations (I count ~178!), generally representing a willingness to engage with
a broad spectrum of the adjacent scientific community.

The manuscript covers many features of the 1.0 version of msprime including at
least (my list and organization):

Software / API / innovations:
Clear separations of ancestry and mutation simulations
Improved methods for downstream analyses
Integration with other software including other simulators
Estimating run times of sim\_ancestry
ARG recording
Specific API for specifying demographic histories
Software development model

Implementations of external methods:
Selective sweeps
Instantaneous bottlenecks
Multiple merger coalescents
Discrete time Wright-Fisher
Gene conversion
Mutation models

As well as benchmarks against other software implementations.

The manuscript is well written and comprehensive - I do not find any major
deficiencies. This manuscript handles a wide range of topics with a high degree
of clarity. And I find the manuscript certainly reaches the criteria "... tools
of interest to a wide range of geneticists" and so is a good fit as an
investigation within the "Methods, Technology, \& Resources" subject.

However, I do think it would be possible to increase the clarity and usefulness
of the manuscript with some relatively small changes in organization.
\end{point}

\begin{reply}
Thanks Reviewer 1, TODO
\end{reply}

\begin{point}
As, is, the manuscript lacks a simple overview of the new capabilities of
version 1.0. Instead these are described in separate sections. The manuscript
might benefit from a short summary and or table of the software's basic
capabilities, API innovations, new functionality, etc. This may also provide a
way to structure the larger flow of the paper, which currently seems to touch
on the different topics in a somewhat random order.
\end{point}

\begin{reply}
TODO Issue 207
\end{reply}

\begin{point}
Generally, I found the figures and examples appropriate and well done. However,
I think they could benefit from more cohesive presentations and descriptions. I
will provide a number of examples - figure 3 and figure 4 present running times
of various aspects of msprime, but present similar data in different ways. In
Figure 3, the data points are not presented directly, while they are present on
Figure 4. And based on the descriptions sometimes replicates are summed over
(Figure 7) or averaged over (Figure 8?) to determine the values on the y-axis.
Figure 3 clearly specifies sim\_mutations as the input for the benchmark, while
for most other figures (e.g. 4) it frequently just says something like "running
time for msprime." Sometimes only population-size scaled rates are provided in
descriptions (e.g Figure 5 and 7) while in others both scaled and absolute
sizes are provided. Sometimes "samples" are described to be diploid (figure 1)
or sometime haploid (Figure 3), but other times the ploidy is left implied (e.g
Figure 4). Sometimes the population size is referred to "effective population
size" (e.g Figure 4) and other times just "population size" (Figure 3). More
consistency in these simulations, descriptions, and figures could add clarity
to the presentation. I did find the lack of a disk storage benchmark an
interesting choice, as this aspect of the software is highlighted in the
introduction but is not present in the manuscript. The algorithmic details of
the simulations and methods are generally not included in the current
manuscript (with some presented in the supplement, and some available in the
2016 paper), and are left for citations. I think this decision serves the
manuscript well otherwise it would likely become unwieldy.
\end{point}

\begin{reply}
TODO Issue 202
\end{reply}

\begin{point}
Figure comments
Figure 2
I think it could be made more clear that the genotype matrix is not part of the set of tables.
\end{point}

\begin{reply}
We agree, and have revised Figure 2 and the caption to clarify the distinction between
stored and derived data.
\end{reply}

\begin{point}
Figure 4
It is not clear if these two plots present all the data used for the quadratic
fits, or if these data continue outside of the plots. Samples are described,
but not their ploidy, are they haploids or diploid? From the plots here, it is
not clear if the quadratic relationship holds for sample sizes << 1000. I
suggest this figure should also include a smaller sample size, (eg n=10)
\end{point}

\begin{reply}
TODO
\end{reply}

\begin{point}
Figure 5
I suggest including the absolute Ne and gc rates be included in the legend, in
addition to the scaled rate.
\end{point}

\begin{reply}
TODO
\end{reply}

\begin{point}
Figure 7
Why is an order of magnitude lower recombination rate used here than in other
examples? This choice seems arbitrary and makes it difficult to relate these
benchmarks results to any others in the paper. Samples are diploid or haploid?
I would suggest the absolute Ne and s rates be included in the legend, in
addition to the scaled rate.
\end{point}

\begin{reply}
TODO

The difference is out of necessity to run discoal (note the y scale).
\end{reply}

\begin{point}

Figure 8
Averaged points are referenced but not shown "Each point...". The plot could be
more clear where the data points are and where the lines are interpolating
between data points. Because of how the simulations were described and plotted,
I found it awkward to try to compare the running time of the DTWF and
coalescent models, even if this was not the purpose of the current figure. The
figure says "to ensure we are measuring" I would suggest changing the text to
"with a goal of measuring".
\end{point}

\begin{reply}
TODO
\end{reply}

\begin{point}
Individual line comments
Line 142 - what is the "msp" program? Is this a command line version?
\end{point}

\begin{reply}
TODO
\end{reply}

\begin{point}
Line 370 - the text here makes it sound like Kelleher et al (2016) makes a
statement relating eq. 1, but the Hein et al (2004) is not cited in the Keleher
et al (2016).

\end{point}

\begin{reply}
TODO
\end{reply}

\begin{point}
Line 370 - population size(s)
\end{point}

\begin{reply}
TODO
\end{reply}

\begin{point}

Line 420 - Despite the long description, I was not able to follow how the
previous paragraph implies that "work is spread out relatively evenly on the
coalescent time scale"
\end{point}

\begin{reply}
TODO
\end{reply}

\begin{point}

Line 422 - There is no context to what is meant by "large" or "small" here - is
2 or 10 or 10000 "large".
\end{point}

\begin{reply}
TODO
\end{reply}

\begin{point}
Line 606 - something goes wrong with the line numbering here.
\end{point}

\begin{reply}
TODO
\end{reply}

\begin{point}

Line 607 - The equations for the allele frequencies during sweeps seem slightly
out of place as (I assume) these are not novel to msprime. There are many
places where more details could be provided about methods discussed here, but
in other cases the equations are left for the supplement or other papers.
\end{point}

\begin{reply}
TODO
\end{reply}

@@

\reviewersection

\begin{point}
Kelleher and colleagues present the 1.0 version of 'msprime', the leading
coalescent simulator which is built upon a highly memory- and time-efficient
data structure, introduced by the lead author, which supports exact simulation
under the coalescent model with recombination.

In addition to being highly efficient, the software engineering of msprime is
of the highest quality. The program includes extensive unit and validation
tests, and offers an API that simplifies worksflows and reduces the
inefficiency and sources of error associated with large intermediate text
files.

Kelleher has made every effort to engage with the community, with the result
that msprime now support many specialized features that have been contributed
by other developers. This has the important benefit of reducing the chances
that this software will lack support in future.

In short, this is an admirable project, and a shining example of successful
academic software development.
\end{point}

\begin{reply}
Thanks reviewer 2, TODO.
\end{reply}

\begin{point}

The paper itself however is less polished. It is very long (37 pages / 1200
lines without appendix), mostly because it is in places overly detailed, and
seems to err on the side of completeness rather than conciseness or clarity.
For instance, lines 160-177 describe the data structure in some detail, while
this is published information. Line 192 remarks that "storage space is
dramatically reduced", and goes on to make this more precise and point out some
(fairly obvious) advantages in lines 193-202.
As another example, line 301-4
states that "Simulating mutations ... is efficient" and refers to Fig 3 for
evidence. The rest of the section (304-319) details a number of examples shown
in Fig 3 that do not add much to the story. As a third example, in the
recombination section after useful comparisons with other approaches, line 354
states that the proposed algorithm is still quadratic in the recombination
rate. The long section from line 354 to 426 details the algorithmic reasons,
and use various analytic approximations and simulations to support this. This
material is of interest only to a few algorithm developers, and I would imagine
it would be better placed in an appendix, with the key observations summarized
in the main text.

\end{point}

\begin{reply}
TODO.
\end{reply}

\begin{point}
Another feature of the current paper is that a number of topics return several
times with minor variations. This is particularly true for algorithmic
efficiency, and the API, both of which are discussed multiple times. It is of
course true that algorithmic efficiency is the major distinguishing feature of
msprime; nevertheless, the focus of the current paper seems to be the rich
feature set, and the software development model, as well as the various APIs
that allow for a more integrated and less error prone workflow. I would suggest
to introduce separate sections for algorithmic efficiency and the API, and
discuss the various features (as you do now) in their own separate section, but
focus on these features from a user's rather than a developer's perspective.

\end{point}
\begin{reply}
TODO.
\end{reply}

\begin{point}
Occasionally the paper reads more like documentation than a paper, such as when
the Demography class is introduced and several associated functions are
mentioned - this is not relevant for a paper in my opinion.

\end{point}
\begin{reply}
TODO.
\end{reply}

\begin{point}
The section about the development model is interesting - very few papers
include such a section, and indeed I think this is one of the stand-out
features of this project that it is so successfully supported by many
contributors. My only suggestion here is to remove the listing of number of
lines of code - this is a very questionable proxy for quality or quantity of
work, and the work is impressive as it is.

\end{point}
\begin{reply}
TODO.
\end{reply}

\begin{point}
This manuscript introduces a simulation tool, msprime version 1.0, and
describes features, additions, and performance of the simulator relative to the
earlier msprime software. I will begin by saying that I believe all researchers
in population genomics who use simulations have appreciated the quality work
that has been previously done in the msprime sphere. The contributions proposed
here clearly improve on a number of features, and that of course has value.
\end{point}
\begin{reply}
TODO.
\end{reply}

\begin{point}
However, I did not find this work to be appropriate for Genetics, and would
rather view it as more appropriate for a journal like Bioinformatics. The
fundamental reason is that this basically appears to perform old
functionalities faster, and the added functionalities are modest. Again, both
are valuable, but this work doesn't address much larger biological /
statistical inference problems that are in need of development in this area,
which could justify publication in a journal like Genetics.
\end{point}
\begin{reply}
TODO.
\end{reply}

\begin{point}

Specifically, the need of the field is really to simulate more realistic models
in a coalescent framework, rather than the slower forward-in-time choice that
stands as the only option presently. For example, the practical use of a tool
that only simulates strictly neutral histories, or histories with the addition
of positive selection, is likely rather limited empirically. By which I mean,
the functional regions of a genome of course experience purifying selection ,
but there is growing appreciation of the fact that this may result in
background selection effects that may be widespread across the genomes of many
species. In order for the strictly neutral demographic simulations to be of
value, one must identify genomic regions that are not only neutral but are also
unaffected by selection at linked sites. In a great many organisms, such
regions may well not exist at all; in terms of commonly studied species, these
regions may only exist in a handful of large, coding-sparse vertebrate and
plant genomes (though, even if these regions exist, they are not necessarily
easy to identify, or sufficiently large to perform neutral demographic
estimation). I note that the manuscript avoids any mention or discussion of
this problem, and avoids any citations that quantify this issue, but the
neglect doesn't alleviate the problem. The authors should really consider,
discuss, and utilize the following work carefully in this regard:

Lohmueller et al. 2011. Natural selection affects multiple aspects of genetic
variation at putatively neutral sites across the human genome. PLOS Genetics.

Comeron. 2014. Background selection as baseline for nucleotide variation across
the Drosophila genome. PLOS Genetics.

Elyashiv et al. 2016. A genomic map of the effects of linked selection in
Drosophila. PLOS Genetics.

Comeron. 2017. Background selection as a null hypothesis in population
genomics: insights and challenges from Drosophila studies. PTRSB.

Pouyet et al. 2018. Background selection and biased gene conversion affect more
than 95\% of the human genome and bias demographic inference. eLife.

Torres et al. 2018. Human demographic history has amplified the effects of
background selection across the genome. PLOS Genetics.

Campos \& Charlesworth. 2019. The effects on neutral variability of recurrent
selective sweeps and background selection. Genetics.

Jensen et al. 2019. The importance of the neutral theory in 1968 and 50 years
on: a response to Kern and Hahn 2018. Evolution.

Castellano et al. 2020. Impact of mutation rate and selection at linked sites
on DNA variation across the genomes of humans and other Homininae. GBE.

Torres et al. 2020. The temporal dynamics of background selection in
nonequilibrium populations. Genetics.

Johri et al. 2021. The impact of purifying and background selection on the
inference of population history: problems and prospects. MBE.

Murphy et al. 2021. Broad-scale variation in human genetic diversity levels is
predicted by purifying selection on coding and non-coding elements. bioRxiv.

\end{point}
\begin{reply}
TODO.
\end{reply}

\begin{point}

Relatedly, for the simulations of positive selection, while it has certainly
been common practice in the field for a few decades, it is rather odd to model
a single positively selected site on a background of all other neutral sites.
Positive selection may act on mutations in functional regions, and functional
regions will be largely shaped by purifying selection. Furthermore, the
Hill-Robertson effects between the modeled positively selected mutation and the
neglected array of negatively selected mutations is often non-trivial. In that
sense, the positive selection model is probably only of empirical relevance in
organisms with something approaching free recombination (to eliminate such
linkage effects), as in perhaps HIV or the like. But of course, these sorts of
organisms tend to have small, coding-dense genomes and experience very strong
selective pressures, thus returning to the first problem mentioned above that
regions free of background selection effects may not exist in the first place.
\end{point}
\begin{reply}
TODO.
\end{reply}

\begin{point}

In summary, I believe this to be a nice software addition that belongs as a
Note in a bioinformatics journal. The software solution that would represent an
important biological advance for Genetics would involve simulating more
realistic models as discussed above (as forward simulators are currently
capable of doing), but at a fast enough rate to be applicable to large-scale
simulation-based genomic inference.
\end{point}

\begin{reply}
TODO
\end{reply}
\end{document}

### 2021-11-15 reviewed-paper.tex
answered comments about the number of lines of code

For large $\rho$, recombination events in trapped ancestral material will
dominate, and so we can use this as a proxy for the overall number
of events in Hudson's algorithm.
\citet[Eq.~5.10]{hein2004gene} gave
\begin{equation}\label{eqn-hsw-bound}
\rho (\rho + 1) \left( \sum_{i=1}^{n-1} \frac{1}{i} \right)^2
\end{equation}
as an upper bound
on the number of recombination events within trapped ancestral material
in Hudson's algorithm for $n$ samples.
% Since the sum is well approximated by $\log{n}$,
% a natural conjecture is that the time complexity of Hudson's
% algorithm is $O(\rho^2 \log^2 n)$.
Fig.~\ref{fig-ancestry-perf} shows the observed run time for
simulations with a variety of population size, chromosome
length and sample sizes, and demonstrates that Eq.~\eqref{eqn-hsw-bound}
correctly predicts the quadratic dependence on $\rho$,
as previously conjectured~\citep[Fig.~2]{kelleher2016efficient}.
We also see that the dependence on $n$ is quite weak,
since increasing sample size 100-fold
only increases run time by a factor of 2 or so. However, the
$\log^2{n}$ factor implied by Eq.~\eqref{eqn-hsw-bound}
(the sum is a harmonic number and can be approximated by $\log{n}$)
is not well supported by observed run times (or numbers of events)
except possibly at very
large values of $\rho$. It therefore appears that a different dependence
on $n$ is required to accurately predict simulation time for a
given $\rho$ and $n$.

@@

\subsection*{Development model}
\texttt{Msprime} has a large number of features, encompassing
the functionality of several more specialised simulators
while maintaining excellent performance.
It is developed by a geographically distributed team of volunteers under an
open source community development model, with a strong emphasis
on code quality, correctness, good documentation, and inclusive development.
As in any large code base,
unit tests play a key role in ensuring that new additions behave
as expected and \msprime\ has an extensive suite.
As of the 1.0.0 release \msprime\ consists of around 13K lines of C and 11K
lines of Python, with suites of 122 C tests (7K lines of code) and
1350 Python tests (22K lines of code). These tests are run
automatically on different operating systems on each pull request
(where a contributor proposes a code change), using standard Continuous
Integration (CI) methodology. Other CI services
check for common errors, code formatting issues, and produce reports on
the level of test coverage for the proposed change.

Unit tests are vital for ensuring software quality and correctness, but
they are usually of little value in assessing the statistical properties
of simulations. To validate the correctness of simulation output
we maintain a suite of statistical tests (as of 1.0.0,
217 validation tests, in 6K lines of code). These consist of running many
replicate simulations to check the properties of the output
against other simulators, and where possible against analytical results.
For example, simulations of complex demography are validated against \ms,
selective sweeps against \discoal, and Wright-Fisher simulations
against forwards in time simulations in \SLiM.
This suite of tests is run before every release, to ensure
that statistical errors have not been introduced.

>>>>

For large $\rho$, recombination events in trapped ancestral material will
dominate, and so we can use this as a proxy for the overall number
of events in Hudson's algorithm.
\citet[Eq.~5.10]{hein2004gene} gave
\begin{equation}\label{eqn-hsw-bound}
\rho (\rho + 1) \left( \sum_{i=1}^{n-1} \frac{1}{i} \right)^2
\end{equation}
as an upper bound
on the number of recombination events within trapped ancestral material
in Hudson's algorithm for $n$ samples.
% Since the sum is well approximated by $\log{n}$,
% a natural conjecture is that the time complexity of Hudson's
% algorithm is $O(\rho^2 \log^2 n)$.
Fig.~\ref{fig-ancestry-perf} shows the observed run time for
simulations with a variety of population sizes, chromosome
length and sample sizes, and demonstrates that Eq.~\eqref{eqn-hsw-bound}
correctly predicts the quadratic dependence on $\rho$,
as previously conjectured~\citep[Fig.~2]{kelleher2016efficient}.
We also see that the dependence on $n$ is quite weak,
since increasing sample size 100-fold
only increases run time by a factor of 2 or so. However, the
$\log^2{n}$ factor implied by Eq.~\eqref{eqn-hsw-bound}
(the sum is a harmonic number and can be approximated by $\log{n}$)
is not well supported by observed run times (or numbers of events)
except possibly at very
large values of $\rho$. It therefore appears that a different dependence
on $n$ is required to accurately predict simulation time for a
given $\rho$ and $n$.

@@

\subsection*{Development model}
\texttt{Msprime} has a large number of features, encompassing
the functionality of several more specialised simulators
while maintaining excellent performance.
It is developed by a geographically distributed team of volunteers under an
open source community development model, with a strong emphasis
on code quality, correctness, good documentation, and inclusive development.
As in any large code base,
unit tests play a key role in ensuring that new additions behave
as expected and \msprime\ has an extensive suite.
These tests are run automatically on different operating systems on each pull request
(where a contributor proposes a code change), using standard Continuous
Integration (CI) methodology. Other CI services
check for common errors, code formatting issues, and produce reports on
the level of test coverage for the proposed change.

Unit tests are vital for ensuring software quality and correctness, but
they are usually of little value in assessing the statistical properties
of simulations. To validate the correctness of simulation output
we maintain a suite of statistical tests (as of 1.0.0,
217 validation tests). These consist of running many
replicate simulations to check the properties of the output
against other simulators, and where possible against analytical results.
For example, simulations of complex demography are validated against \ms,
selective sweeps against \discoal, and Wright-Fisher simulations
against forwards in time simulations in \SLiM.
This suite of tests is run before every release, to ensure
that statistical errors have not been introduced.
